---
title: "Data Science Project"
author: "Sai Myint, Yoojin Oh, Liz Ward, Karl Wirth"
date: '2023-03-26'
output: html_document
---

```{r}
#Set working directory and load packages
# setwd('/Users/karlwirth/Documents/GitHub/ECON6375_DSProject') # Change this to your directory

#Add any packages you need to this chunk
library(tidyverse)
library(table1) #for summary stats tables
library(data.table)
library(ggplot2) #for charts
library(stringr)
library(rlang)
```

#### 2) Reading in the data (Part 1 was to download the data)

```{r}
data2020 <- read.csv('/Users/karlwirth/Downloads/2020_lar_csv.csv') #Don't need to run these
data2021 <- read.csv('/Users/karlwirth/Downloads/2021_public_lar.csv') #This one too
```

#### 3) Filtering by state, appending files, reformatting FIPS and loan amount

```{r}
#Filtering by state_code "OH"
OH_data2020 <- filter(data2020, state_code=='OH')
OH_data2021 <- filter(data2021, state_code=='OH')

#Appending the two datasets together
combined_df <- rbind(OH_data2020, OH_data2021)

#Formatting county code to only be 3 digits
combined_df$county_code <- combined_df$county_code - 39000

combined_df$county_code <- ifelse(combined_df$county_code < 100, 
                                  paste0('0', combined_df$county_code), combined_df$county_code)

#Adding comma thousands separator
combined_df$loan_amount <- format(combined_df$loan_amount, big.mark=',', scientific=FALSE)

save(combined_df, file='combined_df.Rdata')
load('combined_df.Rdata') #Start with loading this file and run the code below to make sure the proper variables get added
```

#### 4) Providing statistical summaries:

##### a) a. Provide a table of minimum, maximum, and average loan value by loan type.

```{r}
combined_df$property_value <- as.numeric(combined_df$property_value) #Converting property
#value variable to a number in order to summarize (from a character)

#Creating new loan_type_word variable to translate numeric loan types into proper acronyms
combined_df$loan_type_word <- ifelse(combined_df$loan_type==1, 'Conventional',
                                     ifelse(combined_df$loan_type==2, 'FHA',
                                            ifelse(combined_df$loan_type==3, 'VA',
                                                   ifelse(combined_df$loan_type==4, 'FSA/RHS', 'fail'))))

combined_df %>%
  group_by(loan_type_word) %>% 
  summarize(mean = mean(property_value, na.rm=TRUE), #Had to remove NA's to summarize or else
            min = min(property_value,  na.rm=TRUE),  #it would not run
            max = max(property_value,  na.rm=TRUE))



```

##### b) Provide similar information graphically. Provide a grouped bar chart with the average loan value by loan type grouped by year. Label the averages for each bar on the chart.

```{r}
combined_df$activity_year <- as.character(combined_df$activity_year) #changing class of activity year to a character and not a number

avg_tbl <- combined_df %>%
  group_by(loan_type_word, activity_year) %>% 
  summarize(mean = mean(property_value, na.rm=TRUE), #Had to remove NA's to summarize or else
            min = min(property_value,  na.rm=TRUE),  #it would not run
            max = max(property_value,  na.rm=TRUE)) #NOTE: Summarize function, with a z, creates a summary stats table


ggplot(avg_tbl, aes(x=loan_type_word, y=mean, fill=activity_year)) + 
  geom_bar(stat="identity", position=position_dodge()) +
  scale_y_continuous(limits=c(0,400000), labels = scales::comma) +
  scale_y_continuous(labels = scales::comma) +
  geom_text(position = position_dodge(width= 1), aes(y=mean + 15000, fill=activity_year, label=round(mean), hjust=0.5), angle=0) +
  labs(title="Average Loan Value Mean by Loan Type",
        x ="Loan Type", y = "Mean Loan Value", fill='Year')
  
```

##### c) c. Provide a table of the average value of rate_spread for 2021 and 2020 by property type.

```{r}
combined_df$property_type <- ifelse(combined_df$occupancy_type==1, 'Principal Residence',
                                    ifelse(combined_df$occupancy_type==2, 'Second Residence',
                                           ifelse(combined_df$occupancy_type==3, 'Investment Property', 'fail')))

combined_df$rate_spread <- as.numeric(combined_df$rate_spread) 

prop_val_avg <- combined_df %>% group_by(activity_year, property_type) %>%
  summarise(mean_rate_spread=mean(rate_spread, na.rm=TRUE)) #NOTE: Summarise function, with an s, aggregates values by mean, sum, etc, and groups by certain variables

print(prop_val_avg)
```

##### d) d. Provide a two-way table of the action taken on the loan by derived_sex.

```{r}
combined_df$action_taken_word <- as.numeric(combined_df$action_taken)

#Recoding the numeric values into words
combined_df$action_taken_word <- ifelse(combined_df$action_taken==1, 'Loan Originated',
                                    ifelse(combined_df$action_taken==2, 'Application Approved but not Accepted',
                                           ifelse(combined_df$action_taken==3, 'Application Denied',
                                                  ifelse(combined_df$action_taken==4, 'Application Withdrawn by Applicant',
                                                         ifelse(combined_df$action_taken==5, 'File Closed for Incompleteness',
                                                                ifelse(combined_df$action_taken==6, 'Purchased Loan',
                                                                       ifelse(combined_df$action_taken==7, 'Preapproval Request Denied',
                                                                              ifelse(combined_df$action_taken==8, 'Preapproval Request Approved but not Accepted', 
                                                                                     'fail'))))))))

act_taken_cnt <- table(combined_df$action_taken_word, combined_df$derived_sex) #This creates the two-way table
print(act_taken_cnt)

```

Examine the "Not Available" category. What percentage of the data falls into this category? Does it the distribution of missing data look random over the different actions taken on loans?

```{r}
addmargins(act_taken_cnt) #This is to give totals to answer the question above

pct_NA <- (237823 / 1613713) * 100
print(pct_NA)

pct_NA_purchased <- (113101 / 148395) * 100
print(pct_NA_purchased)
```

14% of the data falls into the category "Sex Not Available". The distribution of missing data looks relatively random over the different actions taken on loans EXCEPT for "Purchased Loans" where sex data is not available for roughly 76% of those observations. For context, missing sex data accounts for roughly 10% of all the other action categories.

##### e) Provide a two-way table of the action taken on the loan by derived_race.

```{r}
# Shortening the race names so the table looks neater
combined_df$derived_race <- ifelse(combined_df$derived_race=='2 or more minority races', '2+ Minority Races',
                                    ifelse(combined_df$derived_race=='Black or African American', 'Black',
                                           ifelse(combined_df$derived_race=='Free Form Text Only', 'Free Form Text',
                                                  ifelse(combined_df$derived_race=='Native Hawaiian or Other Pacific Islander', 'Pacific Islander',
                                                         ifelse(combined_df$derived_race=='Race Not Available', 'NA', combined_df$derived_race)))))

act_taken_cnt_race <- table(combined_df$action_taken_word, combined_df$derived_race)
print(act_taken_cnt_race)
```

#### 5) Provide the following statistical summaries

##### a) Same as above

##### b) Same as above

##### c) Provide a table of the average value of loans for 2021 and 2020 by property type

```{r}

prop_val_avg_loan_val <- combined_df %>% group_by(activity_year, property_type) %>%
  summarise(mean_loan_value=mean(property_value, na.rm=TRUE))

print(prop_val_avg_loan_val)
```

##### d) Provide a two-way table of the action taken on the loan by preapproval.

```{r}
combined_df$preapproval <- ifelse(combined_df$preapproval==1, 'Preapproval Requested',
                                    ifelse(combined_df$preapproval==2, 'Preapproval Not Requested', 'fail'))

act_taken_cnt_preapproval <- table(combined_df$action_taken_word, combined_df$preapproval)
print(act_taken_cnt_preapproval)

results <- data.frame(act_taken_cnt_preapproval)
```

iii. Choose an appropriate graph to represent the data in this table.

```{r}
results$Var1 = str_wrap(results$Var1, width = 10) #Wrapped the long x-axis labels

ggplot(data = results, aes(x = Var1, y = Freq, fill = Var2)) +
  geom_bar(stat = "identity", position = position_dodge(), alpha = 0.75)  +
  ylim(0,200000) +
 # geom_text(aes(label = Freq), fontface = "bold", vjust = 1.5,
             #position = position_dodge(.9), size = 4) +
  labs(x = "Action Taken", y = "Frequency\n", title = "Actions Taken by Preapproval Status", fill = 'Preapproval Status') +
  theme(plot.title = element_text(hjust = 0.5), 
        axis.title.x = element_text(face="bold", colour="red", size = 12),
        axis.title.y = element_text(face="bold", colour="red", size = 12),
        axis.text.x = element_text(angle=0, size=6),
        legend.title = element_text(face="bold", size = 10))
```

#### 6) Limit the data set to only loan type equal to "Conventional" for the rest of the analysis

```{r}
# Loading R Data File
load("~/Documents/GitHub/ECON6375_DSProject/combined_df.Rdata")
```

```{r}
# Filter data
combined_df_Conventional <- combined_df[combined_df$derived_loan_product_type=="Conventional:First Lien",]
```

#### 7) Create a new data set that is summary statistics by year and county: It should include:

a) Average interest rate per county

b) Percentage of actions taken for loan origination (Loan originated vs all other values)

c) Percentage of debt_to_income ratio that are less than 40%

d) Percentage of loans that are for business or commercial purpose.

e) Percentage of loans that are "investment Property" (category of occupancy types)

```{r}
# Change Column data types
combined_df_Conventional$interest_rate <- as.numeric(combined_df_Conventional$interest_rate)

# Check data type of columns
str(combined_df_Conventional)
```

```{r}
# Group the data by year and county
summary_tbl <- combined_df_Conventional %>% group_by(activity_year,county_code) %>%
  summarise(mean_interest = mean(interest_rate,na.rm = TRUE),
            action_taken_percent = (n_distinct(action_taken)/sum(action_taken))*100,
            debt_to_income_percent = all(debt_to_income_ratio<40)/sum(debt_to_income_ratio),
            .groups = 'drop') %>%
  as.data.frame()
```

```{r}
test <- mean(c(combined_df_Conventional$interest_rate))
```

#### 8) For the count of loan approvals by year and county, do the following

##### a) Reshape the data from long to wide, making each county a single row of data with a column for the 2021 count and a county with the 2020 count

##### b) Create a new variable with the percentage difference between 2021 and 2020

##### c) Print the ten counties with the largest percent changes as percent with 2 decimal places

#### 9) Get county level unemployment data for 2021 & 2020 from BLS
